---
title: "Reproducible Research: Peer Assessment 1"
author: "Bart Decaigny"
date: "Wednesday, October 15, 2014"
output: html_document
keep_md: true
---


## Loading and preprocessing the data

The goal of this assignment is to analyse data from personal movement
activity, in this case, the number of steps taken by an anonymous individual
over the course of two months, measured in five minute intervals.
```{r, echo = FALSE}
## the csv is in a zip file, let's read it from there
activity <- read.csv(unz("activity.zip","activity.csv"))
```

## What is mean total number of steps taken per day?
But before there's anything we can do, we need to load the data, remove NA 
values, clean up, aggregate... to be able to show an overview of the steps 
taken per day:  
```{r, echo = FALSE}
##Cleaning out all NA values
complete <- complete.cases(activity)
data <- activity[complete, ]
##summarize per date..
aggregated <- aggregate(data$steps, list(data$date), sum)
## .. and give summarised data user-friendly names
colnames(aggregated) <- c("date","steps")
## not sure if needed: reprocessing date field
data$date <- strptime(data$date, "%Y-%m-%d")
hist(aggregated$steps, 
     main="Total number of steps taken per day", 
     xlab="Total steps per day")
```

The mean and average of steps taken per day can also be calculated:  
```{r, echo=F}
summary(aggregated$steps)
```


## What is the average daily activity pattern?
Instead of summarizig per day, we could also calculate the average number of 
steps taken per 5-minutes interval, and plot it out:
```{r, echo=FALSE}
interval <- aggregate(data$steps, list(data$interval), mean)
colnames(interval) <- c("interval","avg_steps")
plot(interval$interval, interval$avg_steps, type="l", 
     xlab="interval (as hhmm)", 
     ylab="Average number of steps")
```

We can see that the max number of steps taken is 206, at 8:35 AM, when ordering
the intervals by average steps (only top 5 shown):
```{r, echo=FALSE}
head(interval[order(interval$avg_steps,decreasing=TRUE),],5)
```


## Imputing missing values
```{r, echo=FALSE}
total_NA <- summary(activity[is.na(activity$steps),]$steps)["NA's"]
row_tot <- nrow(activity)
```
As mentioned earlier, these calculations use the cleaned up data set, i.e. with
the NA rows removed. In the original table, there are  `r total_NA` NA rows on 
a total of `r row_tot` rows. In stead of working with reduced data set as with 
did thusfar, we could also choose to replace the missing values with calculated 
data, e.g. the average number of steps for that time interval, or for that date.
As we see quite a strong pattern in the average steps per interval, we prefer to
replace the NA values with the interval means.

Let's check what the impact is on the total number of steps per day resp. on 
the average steps per day.  
```{r, echo=FALSE}
merged <- merge(activity, interval, by.x="interval", by.y="interval")
merged[is.na(merged$steps),]$steps <- round(merged[is.na(merged$steps),]$avg_steps,0)
merged$avg_steps <- NULL
aggregated2 <- aggregate(merged$steps, list(merged$date), sum)
colnames(aggregated2) <- c("date","steps")
hist(aggregated2$steps, 
     main="Total number of steps taken per day", 
     xlab="Total steps per day")
```

The mean and average of steps taken per day can also be recalculated:  
```{r, echo=F}
summary(aggregated2$steps)
```
We can conclude that replacing the NAs with interval means has an impact on 
the quantile boundaries, but not on the mean or the median. When we recreate a 
plot for the average number of plots per interval, we see there is no impact either.  
```{r, echo=FALSE}
interval2 <- aggregate(merged$steps, list(merged$interval), mean)
colnames(interval2) <- c("interval","avg_steps")
plot(interval2$interval, interval2$avg_steps, 
     type="l", 
     xlab="interval (as hhmm)", 
     ylab="Average number of steps")
```

This was expected, of course, as we replaced the missing values with the averages 
already used in the plot.


## Are there differences in activity patterns between weekdays and weekends?
A last question we can ask ourselves, is wether there is a differences between steps taken on weekdays versus weekend. Let's put this in a plot.
```{r, echo=FALSE, results='hide'}
Sys.setlocale("LC_TIME","C")
```
```{r, echo=FALSE, results='markup', fig.height=8}
merged$dow <- weekdays(as.Date(merged$date))
saturday <- merged$dow == "Saturday"
sunday <- merged$dow == "Sunday"
weekend <- saturday + sunday
merged$weekend <- factor(weekend)
par(mfrow=c(2,1))
with(subset(merged, weekend == "1"), plot( interval, steps,type="l", main="weekend"))
with(subset(merged, weekend == "0"), plot(interval, steps,type="l", main="weekdays"))
```  
  
  
  As could be expected, the activity kicks in quite a few hours later in the weekend, and peaks in the late afternoon, instead of in the morning.